---
title: "Reproducible Research Peer Assessment 1"
author: "Brian Ostrovsky"
date: "December 4, 2014"
output: html_document
---

## Loading and preprocessing data
The Activity Monitoring Dataset was downloaded from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip on December 4th at 9:56am PST

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken
 

```{r}
activity <- read.csv("~/Desktop/Reproducable Research/activity.csv")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is the mean total number of steps taken per day?

```{r, echo=TRUE}
totalSteps  <- aggregate(steps~date, data = activity, sum, na.rm = TRUE)
par(mar=c(6,9,4,1), cex=.5, las=2)
histogram<-barplot(totalSteps$steps, names.arg = totalSteps$date,ylab="          Steps", main="Number of Steps Per Day", col="red")
```

```{r, echo=TRUE}
meanSteps  <-  mean(totalSteps$steps)
medianSteps <-  median(totalSteps$steps)
```

The mean number of steps is `r as.integer(meanSteps)` and the median is `r medianSteps`.

## What is the average daily activity pattern?

This plot averages the intevals across each of the days in the data set to get to an average of daily activity

```{r, echo=TRUE}
aveInterval <- aggregate(steps ~ interval, activity, mean)
plot(aveInterval, type="l", xlab="Interval", ylab="Average Steps", main="Average Steps Per 5 minute Interval", col="red")
```

Calculating the maximum interval of the average day:

```{r, echo=TRUE}
aveInterval$interval[which.max(aveInterval$steps)]
```



# How many cases are incomplete?

```{r, echo=TRUE}
incompleteCases  <- sum(!complete.cases(activity))
```
There are `r incompleteCases` observations that are incomplete.

I completed the dataset by filling NA values with the average value for that 5 minute segment. The updated dataset is called "fillInNAs."  
```{r, echo=TRUE}
fillInNAs <- numeric()
for (i in 1:nrow(activity)) {
        observation <- activity[i, ]
        if (is.na(observation$steps)) {
                steps <- subset(aveInterval, interval == observation$interval)$steps
        } else {
                steps <- observation$steps
        }
        fillInNAs <- c(fillInNAs, steps)
}

filledActivity <- activity
filledActivity$steps <- fillInNAs
```


## Create another Histogram of the number of steps per day with the dataset after filling in the NAs

```{r, echo=TRUE}
totalStepsNoNA  <- aggregate(steps~date, data = filledActivity, sum, na.rm = TRUE)
par(mar=c(6,9,4,1))
par(cex=.5)
histogram<-barplot(totalStepsNoNA$steps, names.arg = totalStepsNoNA$date,ylab="          Steps", main="Number of Steps Per Day", col="blue")
```

## Calculate Mean and Median Number of Steps With Missing Values Filled In

```{r, echo=TRUE}
meanStepsNoNA  <-  mean(totalStepsNoNA$steps)
medianStepsNoNA <-  as.integer(median(totalStepsNoNA$steps))
```

The mean number of steps is `r as.integer(meanStepsNoNA)` and the median is `r medianStepsNoNA` for the dataset with the missing values replaced with the mean value for that average segment as compared to a mean of `r as.integer(meanSteps)` and a median of `r medianSteps`. Therefore there is no material difference between the revised dataset and the original.

## Make New Dataset for Weekdays

This adds a column for called "weekDayOrEnd" and assigns either weekend or weekday to that field for each row based on the day of the week that observation took place.
```{r, echo=TRUE}
filledActivity$weekday  <- weekdays(filledActivity$date)

for (i in 1:nrow(filledActivity)) {
        if (filledActivity$weekday[i] == "Saturday" | filledActivity$weekday[i] == "Sunday") {
                filledActivity$weekDayOrEnd[i]  <- "weekend" }    
        else {
                filledActivity$weekDayOrEnd[i]  <- "weekday"}
}
```

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
aveIntervalF <- aggregate(steps ~ interval + weekDayOrEnd, data = filledActivity, mean)
names(aveIntervalF) <- c("interval", "weekDayOrEnd", "steps")
library(lattice)
xyplot(steps ~ interval | weekDayOrEnd, aveIntervalF, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Number of steps")
```
